---
title: "PS270 Final Project"
author: "Natalie Thuente"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---


##10/13 Checkpoint 2:

Research Question: Is one political party more likely to post inaccurate or misleading data on social media than the other?
My hypothesis is that the Republican party is more likely to curate false social media content. My behavioral explanation of this is rooted in recent Republican political events and the overall negatively charged hostility since the 2020 election. The Republican party has been outspokenly discontent with the results of the 2020 election, culminating in the January 6th insurrection. These feelings of hostility have been expressed through many social media posts, and I believe that these extreme views in an abundance of emotional content could explain a lack of accuracy in Republican social media posts.
Explanatory variable of interest: Partisan affiliation of the social media post, measured by the partisan identity of the account creating the post
Response variable: Accuracy of post, this is measured by raters at Buzzfeed, given the same criteria for categorization and randomized posts.
To disprove my hypothesis, the amount of inaccurate posts created by Democratic accounts would have to be greater than that of Republican accounts

data from: https://www.kaggle.com/datasets/mrisdal/fact-checking-facebook-politics-pages/data

```{r}
library(tidyverse)
facebook <- read_csv("facebook-fact-check.csv")
facebook
```

##11/3 Checkpoint 3:

This bar plot shows the relationship between Party and accuracy of social media posts. For each accuracy category (4 being the most accurate, 1 being the least), the plot displays the number of posts that fall under this category for each party. Below is a brief description of the criteria used to classify these posts.

Truth Categories:

(4) Mostly True: The post and any related link or image are based on factual information and portray it accurately. This lets them interpret the event/info in their own way, so long as they do not misrepresent events, numbers, quotes, reactions, etc., or make information up. This rating does not allow for unsupported speculation or claims.

(3) Mixture of True and False: Some elements of the information are factually accurate, but some elements or claims are not. This rating should be used when speculation or unfounded claims are mixed with real events, numbers, quotes, etc., or when the headline of the link being shared makes a false claim but the text of the story is largely accurate. It should also only be used when the unsupported or false information is roughly equal to the accurate information in the post or link. Finally, use this rating for news articles that are based on unconfirmed information.

(2) Mostly False: Most or all of the information in the post or in the link being shared is inaccurate. This should also be used when the central claim being made is false.

(1) No Factual Content: This rating is used for posts that are pure opinion, comics, satire, or any other posts that do not make a factual claim. This is also the category to use for posts that are of the “Like this if you think…” variety.

```{r}
parties <- facebook |>
  filter(Category != "mainstream") |>
  mutate(Party = if_else(Category == "left", "Democrat", "Republican"),
         accuracy_rating = case_when(
           Rating == "mostly true" ~ 4,
           Rating == "mixture of true and false" ~ 3,
           Rating == "mostly false" ~ 2,
           Rating == "no factual content" ~ 1)) |>
  group_by(Party, accuracy_rating) |>
  summarize(n = n())
parties
accuracy_plot <- parties |>
  ggplot(aes(x = accuracy_rating, y = n)) +
  geom_bar(stat = "identity", aes(fill = Party), position = position_dodge()) +
  scale_fill_manual(values = c("Democrat" = "steelblue1", "Republican" = "indianred1")) +
  labs(title = "Social Media Post Accuracy per Party", x = "Accuracy Rating", y = "Number of posts", caption = "4: Mostly True
       3: Mixture of True and False
       2: Mostly False
       1: No Factual Content")
accuracy_plot
```



